---
title: 'Chapter 7: OO field guide'
author: "jharner"
date: "June 7, 2015"
output: html_document
---

R has three object oriented systems (plus the base types).

Central to any object-oriented system are the concepts of **class** and **method**. A class defines the behaviour of objects by describing their attributes and their relationship to other classes. The class is also used when selecting methods, functions that behave differently depending on the class of their input. Classes are usually organised in a hierarchy: if a method does not exist for a child, then the parent’s method is used instead; the child **inherits** behaviour from the parent.

R’s three OO systems differ in how classes and methods are defined:  

* S3 implements a style of OO programming called generic-function OO. This is different from most programming languages, like Java, C++, and C#, which implement message-passing OO. With message-passing, messages (methods) are sent to objects and the object determines which function to call. Typically, this object has a special appearance in the method call, usually appearing before the name of the method/message: e.g., `canvas.drawRect("blue")`. S3 is different. While computations are still carried out via methods, a special type of function called a generic function decides which method to call, e.g., `drawRect(canvas, "blue")`. S3 is a very casual system. It has no formal definition of classes.

* S4 works similarly to S3, but is more formal. There are two major differences to S3. S4 has formal class definitions, which describe the representation and inheritance for each class, and has special helper functions for defining generics and methods. S4 also has multiple dispatch, which means that generic functions can pick methods based on the class of any number of arguments, not just one.

* Reference classes, called RC for short, are quite different from S3 and S4. RC implements message-passing OO, so methods belong to classes, not functions. `$` is used to separate objects and methods, so method calls look like `canvas$drawRect("blue")`. RC objects are also  **mutable**: they don’t use R’s usual copy-on-modify semantics, but are modified in place. This makes them harder to reason about, but allows them to solve problems that are difficult to solve with S3 or S4.

There’s also one other system that’s not quite OO, but it’s important to mention here:  

* base types, the internal C-level types that underlie the other OO systems. Base types are mostly manipulated using C code, but they’re important to know about because they provide the building blocks for the other OO systems.

## 7.1 Base types

Underlying every R object is a C structure (or struct) that describes how that object is stored in memory. The struct includes the contents of the object, the information needed for memory management, and, most importantly for this section, a **type**. This is the **base type** of an R object. Base types are not really an object system because only the R core team can create new types.

```{r}
# The type of a function is "closure"
f <- function() {}
typeof(f)
is.function(f)

# The type of a primitive function is "builtin"
typeof(sum)
is.primitive(sum)
```

You may have heard of `mode()` and `storage.mode()`. Ignore these functions because they’re just aliases of the names returned by `typeof()`, and exist solely for S compatibility.

Functions that behave differently for different base types are almost always written in C, where dispatch occurs using switch statements (e.g., `switch(TYPEOF(x)))`. Even if you never write C code, it’s important to understand base types because everything else is built on top of them: S3 objects can be built on top of any base type, S4 objects use a special base type, and RC objects are a combination of S4 and environments (another base type). 

## 7.2 S3

S3 is R’s first and simplest OO system. It is the only OO system used in the base and stats packages, and it’s the most commonly used system in CRAN packages. S3 is informal and ad hoc, but it has a certain elegance in its minimalism: you can’t take away any part of it and still have a useful OO system.

### 7.2.1 Recognising objects, generic functions, and methods

Most objects that you encounter are S3 objects. But unfortunately there’s no simple way to test if an object is an S3 object in base R. The closest you can come is is.object(x) & !isS4(x), i.e., it’s an object, but not S4. An easier way is to use `pryr::otype()`:
```{r}
library(pryr)

df <- data.frame(x = 1:10, y = letters[1:10])
otype(df)    # A data frame is an S3 class
otype(df$x)  # A numeric vector isn't
otype(df$y)  # A factor is
```

In S3, methods belong to functions, called generic functions, or generics for short. S3 methods do not belong to objects or classes. This is different from most other programming languages, but is a legitimate OO style.

To determine if a function is an S3 generic, you can inspect its source code for a call to `UseMethod()`: that’s the function that figures out the correct method to call, the process of method dispatch. Similar to `otype()`, pryr also provides `ftype()` which describes the object system, if any, associated with a function:
```{r}
mean
ftype(mean)
```

Some S3 generics, like `[`, `sum()`, and `cbind()`, don’t call `UseMethod()` because they are implemented in C. Instead, they call the C functions `DispatchGroup()` or `DispatchOrEval()`. Functions that do method dispatch in C code are called internal generics and are documented in `?"internal generic"`. `ftype()` knows about these special cases too.

Given a class, the job of an S3 generic is to call the right S3 method. You can recognise S3 methods by their names, which look like `generic.class()`. For example, the Date method for the `mean()` generic is called `mean.Date()`, and the factor method for `print()` is called `print.factor()`.

This is the reason that most modern style guides discourage the use of `.` in function names: it makes them look like S3 methods. For example, is `t.test()` the test method for t objects? Similarly, the use of `.` in class names can also be confusing: is `print.data.frame()` the `print()` method for data.frames, or the `print.data()` method for frames? `pryr::ftype()` knows about these exceptions, so you can use it to figure out if a function is an S3 method or generic:
```{r}
ftype(t.data.frame) # data frame method for t()
ftype(t.test)       # generic function for t tests
```

You can see all the methods that belong to a generic with methods():
```{r}
methods("mean")
methods("t.test")
```

You can also list all generics that have a method for a given class:
```{r}
methods(class = "ts")
```

### 7.2.2 Defining classes and creating objects

S3 is a simple and ad hoc system; it has no formal definition of a class. To make an object an instance of a class, you just take an existing base object and set the class attribute. You can do that during creation with `structure()`, or after the fact with `class<-()`:
```{r}
# Create and assign class in one step
foo <- structure(list(), class = "foo")

# Create, then set class
foo <- list()
class(foo) <- "foo"
```

S3 objects are usually built on top of lists, or atomic vectors with attributes. (You can refresh your memory of attributes with attributes.) You can also turn functions into S3 objects. Other base types are either rarely seen in R, or have unusual semantics that don’t work well with attributes.

You can determine the class of any object using `class(x)`, and see if an object inherits from a specific class using `inherits(x, "classname")`.
```{r}
class(foo)
inherits(foo, "foo")
```

The class of an S3 object can be a vector, which describes behaviour from most to least specific. For example, the class of the `glm()` object is `c("glm", "lm")` indicating that generalised linear models inherit behaviour from linear models. Class names are usually lower case, and you should avoid .. Otherwise, opinion is mixed whether to use underscores (my_class) or CamelCase (MyClass) for multi-word class names.

Most S3 classes provide a constructor function:
```{r}
foo <- function(x) {
  if (!is.numeric(x)) stop("X must be numeric")
  structure(list(x), class = "foo")
}
```

You should use it if it’s available (like for `factor()` and `data.frame()`). This ensures that you’re creating the class with the correct components. Constructor functions usually have the same name as the class.

Apart from developer supplied constructor functions, S3 has no checks for correctness. This means you can change the class of existing objects:
```{r}
# Create a linear model
mod <- lm(log(mpg) ~ log(disp), data = mtcars)
class(mod)
print(mod)

# Turn it into a data frame (?!)
class(mod) <- "data.frame"

# But unsurprisingly this doesn't work very well
print(mod)

# However, the data is still there
mod$coefficients
```

If you’ve used other OO languages, this might make you feel queasy. But surprisingly, this flexibility causes few problems: while you can change the type of an object, you never should.

### 7.2.3 Creating new methods and generics

To add a new generic, create a function that calls `UseMethod()`. `UseMethod()` takes two arguments: the name of the generic function, and the argument to use for method dispatch. If you omit the second argument it will dispatch on the first argument to the function. There’s no need to pass any of the arguments of the generic to `UseMethod()` and you shouldn’t do so. `UseMethod()` uses black magic to find them out for itself.
```{r}
f <- function(x) UseMethod("f")
```

A generic isn’t useful without some methods. To add a method, you just create a regular function with the correct (generic.class) name:
```{r}
f.a <- function(x) "Class a"
a <- structure(list(), class = "a")
class(a)
f(a)
```

Adding a method to an existing generic works in the same way:
```{r}
mean.a <- function(x) "a"
mean(a)
```

As you can see, there’s no check to make sure that the method returns the class compatible with the generic. It’s up to you to make sure that your method doesn’t violate the expectations of existing code.

### 7.2.4 Method dispatch

S3 method dispatch is relatively simple. `UseMethod()` creates a vector of function names, like `paste0("generic", ".", c(class(x), "default"))` and looks for each in turn. The “default” class makes it possible to set up a fall back method for otherwise unknown classes.
```{r}
f <- function(x) UseMethod("f")
f.a <- function(x) "Class a"
f.default <- function(x) "Unknown class"

f(structure(list(), class = "a"))

# No method for b class, so uses method for a class
f(structure(list(), class = c("b", "a")))

# No method for c class, so falls back to default
f(structure(list(), class = "c"))
```

Group generic methods add a little more complexity. Group generics make it possible to implement methods for multiple generics with one function. The four group generics and the functions they include are:

* Math: abs, sign, sqrt, floor, cos, sin, log, exp, …  
* Ops: +, -, *, /, ^, %%, %/%, &, |, !, ==, !=, <, <=, >=, >  
* Summary: all, any, sum, prod, min, max, range  
* Complex: Arg, Conj, Im, Mod, Re  

Group generics are a relatively advanced technique and are beyond the scope of this chapter but you can find out more about them in ?groupGeneric. The most important thing to take away from this is to recognise that Math, Ops, Summary, and Complex aren’t real functions, but instead represent groups of functions. Note that inside a group generic function a special variable  .Generic   provides the actual generic function called.

If you have complex class hierarchies it’s sometimes useful to call the “parent” method. It’s a little bit tricky to define exactly what that means, but it’s basically the method that would have been called if the current method did not exist. Again, this is an advanced technique: you can read about it in ?NextMethod.

Because methods are normal R functions, you can call them directly:
```{r}
c <- structure(list(), class = "c")
# Call the correct method:
f.default(c)

# Force R to call the wrong method:
f.a(c)
```

However, this is just as dangerous as changing the class of an object, so you shouldn’t do it. Please don’t point the loaded gun at your foot! The only reason to call the method directly is that sometimes you can get considerable performance improvements by skipping method dispatch. See performance for details.

You can also call an S3 generic with a non-S3 object. Non-internal S3 generics will dispatch on the implicit class of base types. (Internal generics don’t do that for performance reasons.) The rules to determine the implicit class of a base type are somewhat complex, but are shown in the function below:
```{r}
iclass <- function(x) {
  if (is.object(x)) {
    stop("x is not a primitive type", call. = FALSE)
  }

  c(
    if (is.matrix(x)) "matrix",
    if (is.array(x) && !is.matrix(x)) "array",
    if (is.double(x)) "double",
    if (is.integer(x)) "integer",
    mode(x)
  )
}
iclass(matrix(1:5))

iclass(array(1.5))
```



