---
title: 'Chapter 6: Functions'
author: "jharner"
date: "May 31, 2015"
output: html_document
---

Functions are a fundamental building block of R: to master many of the more advanced techniques in this book, you need a solid foundation in how functions work. The most important thing to understand about R is that functions are objects in their own right. You can work with them exactly the same way you work with any other type of object. This theme will be explored in depth in functional programming.

## 6.1 Function components

All R functions have three parts:  

* the `body()`, the code inside the function.  
* the `formals()`, the list of arguments which controls how you can call the function.  
* the `environment()`, the “map” of the location of the function’s variables.

When you print a function in R, it shows you these three important components. If the environment isn’t displayed, it means that the function was created in the global environment.
```{r}
f <- function(x) x^2
f
formals(f)
body(f)
environment(f)
```

The assignment forms of `body()`, `formals()`, and `environment()` can also be used to modify functions.

Like all objects in R, functions can also possess any number of additional `attributes()`. One attribute used by base R is `srcref`, short for source reference, which points to the source code used to create the function. Unlike `body()`, this contains code comments and other formatting. You can also add attributes to a function. For example, you can set the `class()` and add a custom `print()` method.

### 6.1.1 Primitive functions

There is one exception to the rule that functions have three components. Primitive functions, like `sum()`, call C code directly with `.Primitive()` and contain no R code. Therefore their `formals()`, `body()`, and `environment()` are all `NULL`:
```{r}
sum
formals(sum)
body(sum)
environment(sum)
```

Primitive functions are only found in the `base` package, and since they operate at a low level, they can be more efficient (primitive replacement functions don’t have to make copies), and can have different rules for argument matching (e.g., `switch` and `call`).

## 6.2 Lexical scoping

Scoping is the set of rules that govern how R looks up the value of a symbol. In the example below, scoping is the set of rules that R applies to go from the symbol x to its value 10:
```{r}
x <- 10
x
```

Understanding scoping allows you to:  

* build tools by composing functions, as described in functional programming.
* overrule the usual evaluation rules and do non-standard evaluation.

R has two types of scoping: **lexical scoping*, implemented automatically at the language level, and **dynamic scoping**, used in select functions to save typing during interactive analysis. We discuss lexical scoping here because it is intimately tied to function creation. Dynamic scoping is described in more detail in scoping issues.

Lexical scoping looks up symbol values based on how functions were nested when they were created, not how they are nested when they are called. With lexical scoping, you don’t need to know how the function is called to figure out where the value of a variable will be looked up. You just need to look at the function’s definition.

There are four basic principles behind R’s implementation of lexical scoping:  
* name masking  
* functions vs. variables  
* a fresh start  
* dynamic lookup  

### 6.2.1 Name masking

The following example illustrates the most basic principle of lexical scoping:
```{r}
f <- function() {
  x <- 1
  y <- 2
  c(x, y)
}
f()
rm(f)
```

If a name isn’t defined inside a function, R will look one level up.
```{r}
x <- 2
g <- function() {
  y <- 1
  c(x, y) }
g()
rm(x, g)
```

The same rules apply if a function is defined inside another function: look inside the current function, then where that function was defined, and so on, all the way up to the global environment, and then on to other loaded packages. 
```{r}
x <- 1
h <- function() {
  y <- 2
  i <- function() {
    z <- 3
    c(x, y, z) }
  i()
}
h()
rm(x, h)
```

The same rules apply to closures, functions created by other functions. The following function, `j()`, returns a function.
```{r}
j <- function(x) {
  y <- 2
  function() {
    c(x, y)
  }
}
k <- j(1)
k()
rm(j, k)
```

How does R know what the value of `y` is after the function has been called? It works because `k` preserves the environment in which it was defined and because the environment includes the value of `y`.

### 6.2.2 Functions vs. variables

Finding functions works exactly the same way as finding variables:  
```{r}
l <- function(x) x + 1
m <- function() {
  l <- function(x) x * 2
  l(10)
}
m()
rm(l, m)
```

If you are using a name in a context where it’s obvious that you want a function (e.g., `f(3)`), R will ignore objects that are not functions while it is searching. In the following example `n` takes on a different value depending on whether R is looking for a function or a variable.
```{r}
n <- function(x) x / 2
o <- function() {
  n <- 10
  n(n)
}
o()
rm(n, o)
```

### 6.2.3 A fresh start

What happens to the values in between invocations of a function? What will happen the first time you run this function? What will happen the second time? (If you haven’t seen `exists()` before: it returns `TRUE` if there’s a variable of that name, otherwise it returns `FALSE`.)
```{r}
j <- function() {
  if (!exists("a")) {
    a <- 1
  } else {
    a <- a + 1
  }
  print(a)
}
j()
rm(j)
```
You might be surprised that it returns the same value, 1, every time. This is because every time a function is called, a new environment is created to host execution. A function has no way to tell what happened the last time it was run; each invocation is completely independent (but see mutable states).

### 6.2.4 Dynamic lookup

Lexical scoping determines where to look for values, not when to look for them. R looks for values when the function is run, not when it’s created. This means that the output of a function can be different depending on objects outside its environment:
```{r}
f <- function() x
x <- 15
f()
x <- 20
f()
```
You generally want to avoid this behaviour because it means the function is no longer self-contained.

One way to detect this problem is the `findGlobals()` function from `codetools`. This function lists all the external dependencies of a function:
```{r}
f <- function() x + 1
codetools::findGlobals(f)
```

Another way to try and solve the problem would be to manually change the environment of the function to the `emptyenv()`, an environment which contains absolutely nothing:
```{r}
environment(f) <- emptyenv()
# f()
#> Error in f(): could not find function "+"
```
This doesn’t work because R relies on lexical scoping to find everything, even the `+` operator. It’s never possible to make a function completely self-contained because you must always rely on functions defined in base R or other packages.

Since all of the standard operators in R are functions, you can override them with your own alternatives.
```{r}
`(` <- function(e1) {
  if (is.numeric(e1) && runif(1) < 0.1) {
    e1 + 1
  } else {
    e1
  }
}
replicate(50, (1 + 2))
rm("(")
```
This will introduce a particularly pernicious bug: 10% of the time, 1 will be added to any numeric calculation inside parentheses. This is another good reason to regularly restart with a clean R session!

## 6.3 Every operation is a function call

## 6.4 Function arguments

### 6.4.1 Calling functions

### 6.4.2 Calling a function given a list of arguments

### 6.4.3 Default and missing arguments

### 6.4.4 Lazy evaluation

### 6.4.5 ...






