Stat 523
--------

### Assignment 3: Functionals

#### Explain your code where appropriate.

Newton's method for finding a root of a differentiable function $f$ takes a guess $y$ and computes hopfully an improved guess as:
$$ y - \frac{f(y)}{Df(y)}$$
where $Df$ denotes the derivative of $f$.

1. Create a function called `newton_search` with four arguments: `f`, `df`, `guess`, `conv` (the convergence criterion).
```{r}
newton_search <- function(f, df, guess, conv=0.001) {
# Put your function body here.
}
```
Hint: Define a local functions (or helper function) to compute the improvement and then test for convengence.

2. Use this function to find the root of $sin(x)$ near 3 using the actual symbolic derivative. The exact answer is $\pi$.
```{r}
# Put your R code here.
```

3. In general you may not be able to compute the derivative exactly. Use the symmetric difference quotient to approxiate the derivative of $f$ at $x$ numerically by the defintion:
$$ Df \approx \frac{f(x + h) - f(x - h)}{2h} $$
for small $h$.

Define a function `make_derivative` with arguments `f` and `h`. The result returned should be a function closure that remembers both `f` and `h`.
```{r}
make_derivative <- function(f, h) {
# Put your function body here.
}
```

4. Find the root of $sin(x)$ near 3 using numerical derivatives. 
```{r}
# Put your R code here.
```

5. The log-likelihood of the gamma distribution with scale parameter 1 can be written as:
$$ (\alpha -1)s -n\log\Gamma(\alpha) $$
where $\alpha$ is the shape parameter and $s = \sum \log X_i$ is the sufficient statistic.

Randomly draw a sample of $n = 30$ with a shape parameter of $\alpha = 4.5$. Using `newton_search` and `make_derivative`, find the maximun likelihood estimate of $\alpha$. Use the moment esitmator of $\alpha$, i.e., $\bar X$ as the intial guess. The log-likelihood function in R is:
```{r}
x <- rgamma(n=30, shape=4.5)
gllik <- function() {
  s <- sum(log(x))
  n <- length(x)
  function(a) {
    (a - 1) * s - n * lgamma(a)
  }
}
# Put more R code here.
```
Hint: You must apply `newton_search` to the first and second derivatives (derived numerically using `make_derivative`) of the log-likelihood. Your answer should be near 4.5.



