Stat 523
--------

### Assignment 3: Functionals
#### Doug Raffle (dcraffle@mix.wvu.edu)

#### Explain your code where appropriate.

Newton's method for finding a root of a differentiable function $f$ takes a guess $y$ and computes hopfully an improved guess as:
$$ y - \frac{f(y)}{Df(y)}$$
where $Df$ denotes the derivative of $f$.

1. Create a function called `newton_search` with four arguments: `f`, `df`, `guess`, `conv` (the convergence criterion).
```{r}
newton_search <- function(f, df, guess = 0, conv=0.001){
    ## force guess & rename
    y <- guess

    ## helper function to compute difference.  
    d.step <- function(y.i){
        y.i - y
    }

    ## do-while loop to iteratively search
    repeat{
        y.i <- y - f(y)/df(y)
        ## check convergence, stop if done, else update
        if(d.step(y.i) < conv){
            break
        } else{
            y <- y.i
        }
    }
    return(y)
}
```
Hint: Define a local functions (or helper function) to compute the improvement and then test for convengence.

2. Use this function to find the root of $sin(x)$ near 3 using the actual symbolic derivative. The exact answer is $\pi$.
```{r}
newton_search(f = sin, df = cos, guess = 3, conv = 0.001)
```

3. In general you may not be able to compute the derivative exactly. Use the symmetric difference quotient to approxiate the derivative of $f$ at $x$ numerically by the defintion:
$$ Df \approx \frac{f(x + h) - f(x - h)}{2h} $$
for small $h$.

Define a function `make_derivative` with arguments `f` and `h`. The result returned should be a function closure that remembers both `f` and `h`.
```{r}
make_derivative <- function(f, h){
    function(x) (f(x + h) - f(x - h)) / (2*h)
}
```

4. Find the root of $sin(x)$ near 3 using numerical derivatives. 
```{r}
d.sin <- make_derivative(f = sin, h = .01)
newton_search(f = sin, df = d.sin, guess = 3)
```

5. The log-likelihood of the gamma distribution with scale parameter 1 can be written as:
$$ (\alpha -1)s -n\log\Gamma(\alpha) $$
where $\alpha$ is the shape parameter and $s = \sum \log X_i$ is the sufficient statistic.

Randomly draw a sample of $n = 30$ with a shape parameter of $\alpha = 4.5$. Using `newton_search` and `make_derivative`, find the maximun likelihood estimate of $\alpha$. Use the moment esitmator of $\alpha$, i.e., $\bar X$ as the intial guess. The log-likelihood function in R is:

```{r}
x <- rgamma(n=30, shape=4.5)
gllik <- function() {
    s <- sum(log(x))
    n <- length(x)
    function(a) {
        (a - 1) * s - n * lgamma(a)
    }
}

## Get the likelihood and derivative functions for our observed data
gllik.x <- gllik()
d1.gllik.x <- make_derivative(gllik.x, 0.01)
d2.gllik.x <- make_derivative(d1.gllik.x, 0.01)

## estimate
newton_search(f = d1.gllik.x, df = d2.gllik.x, guess = mean(x))
```
Hint: You must apply `newton_search` to the first and second derivatives (derived numerically using `make_derivative`) of the log-likelihood. Your answer should be near 4.5.



