## Statistics 523  

### Assignment 3: One-Way S3 Package

The file created from `R CMD build` (or similar approaches using `devtools` or RStudio shortcuts) should be submitted along with this file and its pdf output. I should be able to install the package---perhaps from GitHub.

**Extra credit of up to 10 points**: Use `git` for  version control and publish your results on GitHub. If GitHub is used, the link should be provided.
        	
#### 1. Start development of a package called `oneway` using the RStudio approach, i.e., start with a project. This initial version should be derived from your S3 classes developed for assignment 4 in the file `oneway.R` or similar files. 

```
List the R code here.
```
	
#### 2. Edit the DESCRIPTION file.

```
List the DESCRIPTION file here.
```

#### 3. Generate a NAMESPACE file using `roxygen2`. That is, export (or less likely import) appropriate functions or S3 methods.

```
List the NAMESPACE file here.
```

#### 4. Develop a help file called `oneway.Rd`, or similar files, which contains the help information for all the functions defined in `oneway.R`, or in similar files. The help file(s) should be placed in a subdirectory called `man`. You should use `roxygen2` for this process.

#### 5. Create an image file from the `coagulation` data set using `save()` and place `coagulation.rda` in the subdirectory `data`.
```{r}
library(faraway)
data(coagulation)
# R code here.
```

#### 6. Develop a help file called `coagulation.Rd`, which contains the help information for the `coagulation` data set. This file should be placed in the `man` subdirectory. Again use `roxygen2`.

#### 7. Build the package using the `R CMD build` and the `R CMD check` commands or by RStudio commands or `devtools`. The process should be iterated until bugs are found and corrected. You can use the debugging tools you have learned. Once complete, install the package.
	
#### 8. Show the output from the help files using: `library(oneway)`, `help(oneway)`, help for specific functions, and `help(coagulation)`, etc.
```{r oneway.help}
# library(oneway)
# R code here 
```
The help files should provide all relevant information.

#### 9. Illustrate your functions using:  
```{r oneway.example}
# example(oneway)
```
The example should illustrate all aspects of your code.
