---
title: "Data structures"
author: "jharner"
date: "May 18, 2015"
output: html_document
---

R’s base data structures can be organised by their dimensionality (1d, 2d, or nd) and whether they’re homogeneous (all contents must be of the same type) or heterogeneous (the contents can be of different types). Almost all other objects are built upon these foundations.

Given an object, the best way to understand what data structures it’s composed of is to use `str()`.

## Vectors

The basic data structure in R is the vector. Vectors come in two flavours: atomic vectors and lists. They have three common properties:  
* Type, `typeof()`, what it is.  
* Length, `length()`, how many elements it contains.  
* Attributes, `attributes()`, additional arbitrary metadata.  
All elements of an atomic vector must be the same type, whereas the elements of a list can have different types.

### Atomic vectors

There are four common types of atomic vectors that I’ll discuss in detail: logical, integer, double (often called numeric), and character.

Atomic vectors are usually created with c(), short for combine:
```{r}
dbl_var <- c(1, 2.5, 4.5)
dbl_var

# With the L suffix, you get an integer rather than a double
int_var <- c(1L, 6L, 10L)
int_var

# Use TRUE and FALSE (or T and F) to create logical vectors
log_var <- c(TRUE, FALSE, T, F)
log_var

chr_var <- c("these are", "some strings")
chr_var
```

Atomic vectors are always flat, even if you nest c()’s:
```{r}
c(1, c(2, c(3, 4)))

# the same as
c(1, 2, 3, 4)
```

Missing values are specified with `NA`, which is a logical vector of length 1. `NA` will always be coerced to the correct type if used inside `c()`, or you can create NAs of a specific type with `NA_real_` (a double vector), `NA_integer_` and `NA_character_`.

#### Types and tests

Given a vector, you can determine its type with `typeof()`, or check if it’s a specific type with an “is” function: `is.character()`, `is.double()`, `is.integer()`, `is.logical()`, or, more generally, `is.atomic()`.
```{r}
int_var <- c(1L, 6L, 10L)
typeof(int_var)
is.integer(int_var)
is.atomic(int_var)

dbl_var <- c(1, 2.5, 4.5)
typeof(dbl_var)
is.double(dbl_var)
is.atomic(dbl_var)
```

`is.numeric()` is a general test for the “numberliness” of a vector.
```{r}
is.numeric(int_var)
is.numeric(dbl_var)
```

#### Coercion

All elements of an atomic vector must be the same type, so when you attempt to combine different types they will be coerced to the most flexible type. Types from least to most flexible are: logical, integer, double, and character.

Combining a character and an integer yields a character:
```{r}
str(c("a", 1))
```

When a logical vector is coerced to an integer or double, TRUE becomes 1 and FALSE becomes 0.
```{r}
x <- c(FALSE, FALSE, TRUE)
as.numeric(x)

# Total number of TRUEs
sum(x)

# Proportion that are TRUE
mean(x)
```

Coercion often happens automatically. Most mathematical functions (+, log, abs, etc.) will coerce to a double or integer, and most logical operations (&, |, any, etc) will coerce to a logical.

### Lists

Lists are different from atomic vectors because their elements can be of any type, including lists.
```{r}
x <- list(1:3, "a", c(TRUE, FALSE, TRUE), c(2.3, 5.9))
str(x)
```

Lists are sometimes called recursive vectors.
```{r}
x <- list(list(list(list())))
x
str(x)
is.recursive(x)
```

`c()` will combine several lists into one.
```{r}
x <- list(list(1, 2), c(3, 4))
x
str(x)
y <- c(list(1, 2), c(3, 4))
y
str(y)
```

The `typeof()` a list is list. You can test for a list with `is.list()` and coerce to a list with `as.list()`. You can turn a list into an atomic vector with `unlist()`. If the elements of a list have different types, `unlist()` uses the same coercion rules as `c()`.

Data frames (described in data frames) and linear models objects (as produced by `lm()`) are lists:
```{r}
is.list(mtcars)

mod <- lm(mpg ~ wt, data = mtcars)
is.list(mod)
```

## Attributes

Attributes can be thought of as a named list (with unique names). Attributes can be accessed individually with `attr()` or all at once (as a list) with `attributes()`.
```{r}
y <- 1:10
attr(y, "my_attribute") <- "This is a vector"
attr(y, "my_attribute")
str(attributes(y))
```

The `structure()` function returns a new object with modified attributes:
```{r}
structure(1:10, my_attribute = "This is a vector")
```

By default, most attributes are lost when modifying a vector:
```{r}
attributes(y[1])
attributes(sum(y))
```

The only attributes not lost are the three most important:  
* Names, a character vector giving each element a name  
* Dimensions, used to turn vectors into matrices and arrays  
* Class, used to implement the S3 object system  

Each of these attributes has a specific accessor function to get and set values. When working with these attributes, use `names(x)`, `dim(x)`, and `class(x)`, not `attr(x, "names")`, `attr(x, "dim")`, and `attr(x, "class")`.

#### Names

You can name a vector in three ways:  
* When creating it: x <- c(a = 1, b = 2, c = 3).  
* By modifying an existing vector in place: x <- 1:3; names(x) <- c("a", "b", "c").  
* By creating a modified copy of a vector: x <- setNames(1:3, c("a", "b", "c")).  

Names don’t have to be unique. Not all elements of a vector need to have a name. If some names are missing, `names()` will return an empty string for those elements. If all names are missing, `names()` will return NULL.
```{r}
y <- c(a = 1, 2, 3)
names(y)
z <- c(1, 2, 3)
names(z)
```

You can create a new vector without names using `unname(x)`, or remove names in place with `names(x) <- NULL`.

### Factors

A factor is a vector that can contain only predefined values, and is used to store categorical data. Factors are built on top of integer vectors using two attributes: the `class()`, “factor”, which makes them behave differently from regular integer vectors, and the `levels()`, which defines the set of allowed values.
```{r}
x <- factor(c("a", "b", "b", "a"))
x
class(x)
levels(x)

# You can't use values that are not in the levels
x[2] <- "c"
x

# NB: you can't combine factors
c(factor("a"), factor("b"))
```

Factors are useful when you know the possible values a variable may take, even if you don’t see all values in a given dataset. Using a factor instead of a character vector makes it obvious when some groups contain no observations:
```{r}
sex_char <- c("m", "m", "m")
sex_factor <- factor(sex_char, levels = c("m", "f"))

table(sex_char)
table(sex_factor)
```

Sometimes when a data frame is read directly from a file, a column you’d thought would produce a numeric vector instead produces a factor. This is caused by a non-numeric value in the column, often a missing value encoded in a special way like . or -. To remedy the situation, coerce the vector from a factor to a character vector, and then from a character to a double vector. 
```{r}
# Reading in "text" instead of from a file here:
z <- read.csv(text = "value\n12\n1\n.\n9")
typeof(z$value)
as.double(z$value)

# Oops, that's not right: 3 2 1 4 are the levels of a factor, 
# not the values we read in!
class(z$value)

# We can fix it now:
as.double(as.character(z$value))

# Or change how we read it in:
z <- read.csv(text = "value\n12\n1\n.\n9", na.strings=".")
typeof(z$value)
class(z$value)
z$value
```

Unfortunately, most data loading functions in R automatically convert character vectors to factors. This is suboptimal, because there’s no way for those functions to know the set of all possible levels or their optimal order. Instead, use the argument `stringsAsFactors = FALSE` to suppress this behaviour, and then manually convert character vectors to factors using your knowledge of the data. 

While factors look (and often behave) like character vectors, they are actually integers. Be careful when treating them like strings. Some string methods (like `gsub()` and `grepl()`) will coerce factors to strings, while others (like `nchar()`) will throw an error, and still others (like `c()`) will use the underlying integer values.

## Matrices and arrays

Adding a `dim()` attribute to an atomic vector allows it to behave like a multi-dimensional array. A special case of the array is the matrix, which has two dimensions. Matrices are used commonly as part of the mathematical machinery of statistics.

Matrices and arrays are created with `matrix()` and `array()`, or by using the assignment form of `dim()`:
```{r}
# Two scalar arguments to specify rows and columns
a <- matrix(1:6, ncol = 3, nrow = 2)
a

# One vector argument to describe all dimensions
b <- array(1:12, c(2, 3, 2))
b

# You can also modify an object in place by setting dim()
c <- 1:6
dim(c) <- c(3, 2)
c
dim(c) <- c(2, 3)
c
```

`length()` and `names()` have high-dimensional generalisations:  
* `length()` generalises to `nrow()` and `ncol()` for matrices, and `dim()` for arrays.  
* `names()` generalises to `rownames()` and `colnames()` for matrices, and `dimnames()`, a list of character vectors, for arrays.
```{r}
length(a)
nrow(a)
ncol(a)

rownames(a) <- c("A", "B")
colnames(a) <- c("a", "b", "c")
a

length(b)
dim(b)
dimnames(b) <- list(c("one", "two"), c("a", "b", "c"), c("A", "B"))
b
```

`c()` generalises to `cbind()` and `rbind()` for matrices, and to `abind()` (provided by the `abind` package) for arrays. You can transpose a matrix with `t()`; the generalised equivalent for arrays is `aperm()`.

You can test if an object is a matrix or array using `is.matrix()` and `is.array()`, or by looking at the length of the `dim()`. `as.matrix()` and `as.array()` make it easy to turn an existing vector into a matrix or array.

Vectors are not the only 1-dimensional data structure. You can have matrices with a single row or single column, or arrays with a single dimension. They may print similarly, but will behave differently. Use str() to reveal the differences.

```{r}
str(1:3)                   # 1d vector

str(matrix(1:3, ncol = 1)) # column vector

str(matrix(1:3, nrow = 1)) # row vector

str(array(1:3, 3))         # "array" vector
```

While atomic vectors are most commonly turned into matrices, the dimension attribute can also be set on lists to make list-matrices or list-arrays:
```{r}
l <- list(1:3, "a", TRUE, 1.0)
dim(l) <- c(2, 2)
l
```

## Data frames

A data frame is the most common way of storing data in R. Under the hood, a data frame is a list of equal-length vectors. This makes it a 2-dimensional structure, so it shares properties of both the matrix and the list. This means that a data frame has `names()`, `colnames()`, and `rownames()`, although `names()` and `colnames()` are the same thing. The `length()` of a data frame is the length of the underlying list and so is the same as `ncol()`; `nrow()` gives the number of rows.

### Creation

You create a data frame using `data.frame()`, which takes named vectors as input:
```{r}
df <- data.frame(x = 1:3, y = c("a", "b", "c"))
df
str(df)
```

Beware `data.frame()`’s default behaviour which turns strings into factors. Use `stringAsFactors = FALSE` to suppress this behaviour:
```{r}
df <- data.frame(
  x = 1:3,
  y = c("a", "b", "c"),
  stringsAsFactors = FALSE)
df
str(df)
```

### Testing and coercion

Because a data.frame is an S3 class, its type reflects the underlying vector used to build it: the list. To check if an object is a data frame, use `class()` or test explicitly with `is.data.frame()`:
```{r}
typeof(df)
class(df)
is.data.frame(df)
```

You can coerce an object to a data frame with `as.data.frame()`:  
* A vector will create a one-column data frame.  
* A list will create one column for each element; it’s an error if they’re not all the same length.  
* A matrix will create a data frame with the same number of columns and rows as the matrix.
```

### Combining data frames

You can combine data frames using cbind() and rbind():
```{r}
cbind(df, data.frame(z = 3:1))
rbind(df, data.frame(x = 10, y = "z"))
```

When combining column-wise, the number of rows must match, but row names are ignored. When combining row-wise, both the number and names of columns must match. Use `plyr::rbind.fill()` to combine data frames that don’t have the same columns.

It’s a common mistake to try and create a data frame by `cbind()`ing vectors together. This doesn’t work because `cbind()` will create a matrix unless one of the arguments is already a data frame. Instead use `data.frame()` directly:
```{r}
bad <- data.frame(cbind(a = 1:2, b = c("a", "b")))
str(bad)

good <- data.frame(a = 1:2, b = c("a", "b"),
  stringsAsFactors = FALSE)
str(good)
```

The conversion rules for cbind() are complicated and best avoided by ensuring all inputs are of the same type.

### Special columns

Since a data frame is a list of vectors, it is possible for a data frame to have a column that is a list:
```{r}
df <- data.frame(x = 1:3)
df$y <- list(1:2, 1:3, 1:4)
df
```

However, when a list is given to `data.frame()`, it tries to put each item of the list into its own column, so this fails:
```{r}
# data.frame(x = 1:3, y = list(1:2, 1:3, 1:4))
#> Error in data.frame(1:2, 1:3, 1:4, check.names = FALSE, stringsAsFactors = TRUE): arguments imply differing number of rows: 2, 3, 4
```

A workaround is to use `I()`, which causes `data.frame()` to treat the list as one unit:
```{r}
dfl <- data.frame(x = 1:3, y = I(list(1:2, 1:3, 1:4)))
dfl
str(dfl)
dfl[2, "y"]
```

`I()` adds the `AsIs` class to its input, but this can usually be safely ignored.

Similarly, it’s also possible to have a column of a data frame that’s a matrix or array, as long as the number of rows matches the data frame:
```{r}
dfm <- data.frame(x = 1:3, y = I(matrix(1:9, nrow = 3)))
dfm
str(dfm)
dfm[2, "y"]
```

Use list and array columns with caution: many functions that work with data frames assume that all columns are atomic vectors.








